---
title: "Time"
date: "2022"
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_folding: show  
    highlight: tango
    number_sections: true
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.height = 5, fig.width = 8)
```


```{r}
library(tidyverse)          # ggplot2 + more
library(lubridate)          # date and time functions
library(cowplot)            # arrange multiple plots
library(plotly)             # Animations
library(gapminder)          # data for the Animations
```

## Data
We use bakery sales data provided by [Meteolytix](https://meteolytix.de/) for some anonymous bakery of the Kiel region. Additionally, we use weather data from Deutscher Wetterdienst (DWD).

```{r}
sales <- read_csv("data/sales.csv")        # Bakery sales
weather <- read_csv("data/weather.csv")    # Weather data
df <- sales %>% inner_join(weather, by = "date")
```

**Price question**: Can you identify which `product_type` number (1-6) represents which type of product type (Bread, Cakes, Drinks, Patisserie, Rolls, Snacks), just based on the sales patterns?

## Time and date variables 
The `lubridate` package is helpful for all sorts of data manipulations on date and times. Here, we use the functions `year`, `month` (Jan, Feb, ...), `yday` (1-365), `wday` (Monday, Tuesday, ...) to extract relevant information from the `date` variable. Importantly, the month is not simply stored as a character variable, but as an ordered factor. In this way, months get automatically the natural order in plots.

```{r}
df <- df %>% 
  mutate(
    year = lubridate::year(date), 
    month = lubridate::month(date, label = TRUE, abbr = FALSE),
    dayofyear = lubridate::yday(date),
    dayofweek = lubridate::wday(date, label=TRUE, abbr = FALSE, week_start = 1)
  )
```

See the [lubridate cheatsheet](https://rawgit.com/rstudio/cheatsheets/master/lubridate.pdf) for more information.

# Time Series Plots
## Axis labels

With time series plots, we can use the function `scale_x_date` to format the appearance of the date or time axis. For instance, we can define the distance between breaks or grid lines, or define the labels (2020-12-24, 24. December (Thursday), etc.). Use `help(scale_x_date)` to see the options and `help(strptime`) to see specifically the options for the date label formatting.

```{r}
df %>%
  filter(year == 2018, product_type==2) %>%
  ggplot(aes(date, sales)) +
  geom_line() +
  scale_x_date(
    date_breaks = "2 months",          # Distance between labeled breaks
    date_minor_breaks = "1 month",     # Distance between grid lines
    date_labels = "%a, %d.%m."         # Date labels format 
  )
```

## Arrange vertically

If you have two time series in different units, **do not use dual y-axes**. Instead, one option is to arrange them vertically. To do so, we first create two separate plots and then we arrange them vertically using the `cowplot` package. There are several alternative packages for this step. 

```{r}
p1 <- df %>%
  filter(year == 2018) %>%
  ggplot(aes(date, sales)) + 
  geom_smooth(se=FALSE, fill = "blue") +
  labs(y = "Sales")  +
  theme(axis.ticks.x = element_blank(), 
        axis.text.x = element_blank(), 
        axis.title.x = element_blank())

p2 <- df %>%
  filter(product_type==1, year == 2018) %>%
  ggplot(aes(x = date, y = temperature)) +
  geom_line(color = "red", size = 1.5) +
  labs(y = "Temperature")

cowplot::plot_grid(p1, p2, ncol=1, align = "v")
```


### Compare cyclical patterns
Often we have cyclical patterns within a certain interval: yearly, monthly, daily, etc.

```{r}
p <- df %>%
  filter(product_type==2) %>%
  ggplot(aes(x = date, y = sales)) +
  geom_line()

plotly::ggplotly(p)
```

To better recognize cyclical patterns, and analyze how they have changed over time, we create the following plot. In principle, we could map the `year` to the color aesthetic. But this typically overloads the plot with too much lines. Instead, we just do this for a few selected years. The other years are just plotted as grey colored reference.

```{r}
df2 <- df %>% filter(product_type==2) 
df2_subset <- df2 %>% filter(year %in% c(2014, 2018))
df2 %>%
  ggplot(aes(x = dayofyear, y = sales, group = year)) +
  geom_smooth(color = "grey", se=FALSE) +
  geom_smooth(data = df2_subset, aes(color = as.ordered(year)), se=FALSE) +
  labs(y = "Temperature", color = "Year", y = "Temperture", x = "Day of Year",
       title = "Seasonal sales profiles") 
```


# Beyond Time Series
## Polar plot
Polar plots are generally difficult to read. However, they highlight the cyclical nature of time. Suppose we make a bar plot in a regular, cartesian coordinate system, then there is a disconnect between December and January. By changing to a **polar coordinate system** we highlight the fact that December is followed by January (Sunday is followed by Monday, 23.59h is followed by 00.00h, etc.)

```{r}
df %>%
  group_by(month) %>%
  summarise(temperature = mean(temperature, na.rm=TRUE), 
            precipitation = sum(precipitation,na.rm=TRUE)) %>%
  ggplot(aes(x = month, y = temperature, fill = precipitation)) +
  geom_col() +
  scale_fill_steps(low = "grey", high = "darkblue") + 
  coord_polar(start = pi)

```

## Calendar Plot

We can do even fancy things like a calendar plot, which highlights interesting seasonal, monthly, and weekly patterns as well as outliers.

The code gets complicated, but remember: once we have figured out how to do it, we can wrap the code inside a function, and create nice calendar plots as a one liner.

```{r}
calender_heatmap <- function(data, datevar, colorvar){
  
  # Variables
  data <- data %>%
    mutate(
      dayofweek = wday({{datevar}}, label = TRUE, week_start = 1),   
      month = month({{datevar}}, label = TRUE),                      
      week = isoweek({{datevar}}),                                    
      dayofmonth = mday({{datevar}})) %>%                             
    mutate(week = if_else(month == "Dez" & week==1, 53, week))
  
  # Plot
  data %>%
    ggplot(aes(x = dayofweek, y = week)) +                   
    geom_tile(aes(fill = {{colorvar}}), color = "white") +          
    geom_text(aes(label = dayofmonth), size =2, color = "grey") +   
    facet_wrap(~month, nrow = 3, ncol = 4, scales = "free_y") + 
    scale_y_reverse() +                                         
    scale_fill_viridis_c() +
    theme_bw() +
    theme(                                  # ... with some adjustments 
      axis.title = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank()
    )
}
```

```{r}
p <- df %>% 
  filter(product_type==1, year==2018) %>%
  calender_heatmap(datevar = date, colorvar = sales)

plotly::ggplotly(p)
```


## Animation
Last, but not least, animations are a way of adding a time representation on top of existing plots. Some basic animation functionality is implemented in the `plotly` package, which works well for scatterplots. Here, the animation just requires a further mapping: `year` is mapped to the aesthetic `frame` . Then the ggplot object is passed to `plotly::ggplotly`. 

The problem with animations is that they often carry too much information. You will need to fine tune the animation options to make it nice looking. Animations are sometimes a nice way to draw attention, but they rarely convey information in an effective way.

Note that there is an R package `gganimate` that can add animations on top of arbitrary plots (scatterplots, bar plots, maps, etc). However, since it is based on GIFs rather than HTML the appearance is less modern. 

```{r}
p <- gapminder %>% 
  ggplot(aes(x = gdpPercap,            # Standard aesthetic mapping ...
             y = lifeExp, 
             size = pop, 
             color = continent,
             frame = year,             # ... mapping to animation frame
             ids=country)) +           # ... adds country to tooltip
  geom_point() +  
  scale_x_log10(labels = scales::comma)

plotly::ggplotly(p) %>%                # Turn into plotly plot
  animation_opts(frame = 1000,         # Frame time (milliseconds)
                 transition = 1000
  ) 
```

