---
title: "Problemset: Time"
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_folding: show  
    highlight: tango
    number_sections: true
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


```{r}
library(tidyverse) 
library(lubridate)
library(ggrepel)
theme_set(theme_bw())
```

## Data
We use bakery sales data provided by [Meteolytix](https://meteolytix.de/) for some unknown bakery of the Kiel region. Additionally, we use weather data from Deutscher Wetterdienst (DWD).

```{r}
sales <- read_csv("data/sales.csv")        # Bakery sales
weather <- read_csv("data/weather.csv")    # Weather data
df <- sales %>% left_join(weather, by = "date")
```

## Feature engineering
```{r}
df <- df %>%
  mutate(
    wday = lubridate::wday(date, label=TRUE, abbr=FALSE, week_start = 1),
    month = month(date, label=TRUE, abbr = FALSE),
    year = year(date),
    dayofyear = yday(date))
```

The product types in the data are represented with anonymous numbers (1-6). Part of your task is to infer which number corresponds to which actual type:  bread, rolls, cakes, drinks, patisserie, snacks. 


Visualize the full time series of total daily sales. Which trends or patterns do you see?

```{r}
df %>%
  group_by(date) %>%
  summarise(sales = sum(sales)) %>%
  ggplot(aes(date, sales)) +
  geom_line(size=0.1, alpha=0.6) +
  geom_smooth(formula=y ~ s(x, bs='cs', k=20), size=2) +   # Smoothness paramter: k
  scale_x_date(
    date_breaks = "12 months",        # Distance between labeled breaks
    date_minor_breaks = "3 month",    # Distance between grid lines
    date_labels = "%Y\n%b-%d"         # Date labels format 
  )

```




```{r}
df %>% 
  mutate(week_year = lubridate::floor_date(date, unit="weeks", week_start=1)) %>% 
  group_by(week_year) %>%
  summarise(sales = sum(sales)) %>%
  ggplot(aes(week_year, sales)) +
  geom_line(size=1) + geom_point() +
  scale_x_date(
    date_breaks = "12 months",          # Distance between labeled breaks
    date_minor_breaks = "6 month",     # Distance between grid lines
    date_labels = "%Y-%m-%d"         # Date labels format 
  ) +
  scale_y_continuous(labels=scales::comma)
```


Visualize the full time series of daily sales per product type. Which product-specific patterns do you perceive?
```{r}
df %>% 
  mutate(month_date = lubridate::floor_date(date, unit="months")) %>% 
  group_by(month_date, product_type = factor(product_type)) %>%
  summarise(sales = sum(sales)) %>%
  ggplot(aes(month_date, sales, color = product_type)) +
  geom_line(size = 1) + geom_point(size = 2) +
  #geom_smooth(size=1, span=0.3, se=FALSE) + 
  scale_x_date(
    date_breaks = "12 months",          # Distance between labeled breaks
    date_minor_breaks = "6 month",     # Distance between grid lines
    date_labels = "%Y-%m-%d"         # Date labels format 
  ) +
  scale_y_continuous(labels=scales::comma) +
  scale_color_brewer(palette = "Set1")

```


Zoom in on a particular year to visualize the intra-year patterns. Which patterns do you see?
```{r}

df %>% mutate(holiday = ifelse(month=='Dezember' & day(date)==31, 1, 0))

df %>% 
  filter(year==2018) %>%
  mutate(holiday = ifelse(month=="Dezember" & day(date)==31, "New Year's Eve", NA)) %>%
  group_by(date, holiday) %>% 
  summarise(sales = mean(sales)) %>% 
  ggplot(aes(date, sales)) +
  geom_line() + geom_point() +
  geom_text_repel(aes(label = holiday))

```


Visualize intra-week sales patterns. On which day does the bakery sell the most/least?
```{r}
df %>%
  group_by(wday) %>%
  summarise(sales = mean(sales)) %>%
  ggplot(aes(x=wday, y=sales)) +
  geom_point(size = 3) + geom_line(aes(group = 1)) +
  scale_color_brewer(palette = "Set1")
```
```{r}
df %>%
  group_by(wday) %>%
  summarise(sales = mean(sales)) %>%
  ggplot(aes(x=wday, y=sales)) +
  geom_col(fill = 'lightgrey', col = "black") +
  scale_color_brewer(palette = "Set1")
```



Visualize the average daily sales per weekday and product type? Which product-specific patterns do you see?
```{r}
df %>%
  group_by(wday, product_type = factor(product_type)) %>%
  summarise(sales = mean(sales)) %>%
  ggplot(aes(x=wday, y=sales, color=product_type)) +
  geom_point(size = 3) + 
  geom_line(aes(group=product_type), size = 1.5) +
  scale_color_brewer(palette = "Set1")

```

```{r}
df %>%
  group_by(wday, product_type = factor(product_type)) %>%
  summarise(sales = mean(sales)) %>%
  ggplot(aes(x=wday, y=sales)) +
  geom_col() + 
  scale_color_brewer(palette = "Set1") +
  facet_wrap(vars(product_type)) +
  theme(axis.text.x = element_text(angle = 90))
```


Is there a relationship between weather features (sunshine, temperature, rain) and sales?
```{r}
df %>%
  ggplot(aes(x = temperature, y = sales)) +
  geom_point(alpha = 0.1) + geom_smooth(size = 1.5) +
  facet_wrap(vars(product_type), scales = "free_y")

```

```{r}
df %>%
  ggplot(aes(x = sunhours, y = sales)) +
  geom_point(alpha = 0.1) + geom_smooth(size = 1.5) +
  facet_wrap(vars(product_type), scales = "free_y")
```




